### 2016-02-25 stitched.tex
Final submission.

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
    \setmainfont{Georgia}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

\usepackage{booktabs}
\usepackage[final]{changes}
\usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
\linespread{1.5}
\usepackage[compact]{titlesec}
\usepackage{enumitem}
\usepackage{tikz}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
\setlist{nolistsep}
\titlespacing{\section}{2pt}{*0}{*0}
\titlespacing{\subsection}{2pt}{*0}{*0}
\titlespacing{\subsubsection}{2pt}{*0}{*0}
\setlength{\parskip}{3pt}
\setremarkmarkup{(#2)}
\definechangesauthor[name={Nick Tustison}, color=red]{nt}
\definechangesauthor[name={James Stone}, color=magenta]{js}
\definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
\definechangesauthor[name={Andy Mayer}, color=green]{am}
\definechangesauthor[name={Harvey Levin}, color=blue]{hl}
\definechangesauthor[name={Brian Taylor}, color=purple]{bt}
\definechangesauthor[name={David Tate}, color=orange]{dt}

@@

Machine learning and pattern recognition techniques have seen increased
application for various medical image analysis workflows (see, for
example, the annual Workshop on Machine Learning in Medical Imaging held
in conjunction with the Medical Image Computing and Computer-Aided
Intervention (MICCAI) international meeting\textsuperscript{11}).
Popular techniques such as support vector machines and neural networks
have been applied successfully to clinically relevant imaging tasks such
as supervised image segmentation (e.g.,\textsuperscript{12}) and
diagnostic prediction (e.g.,\textsuperscript{13,14}). Facilitating the
current employment of such techniques are the number of available
imaging data sets\textsuperscript{15} and the public availability of
data science packages such as SciPy\textsuperscript{16} and the R
project for statistical computing\textsuperscript{17} and their
associated extensions.

The random forests framework\textsuperscript{18} is a popular machine
learning technique that has demonstrated significant
\replaced[id=dt]{utility}{utitility} for supervised segmentation tasks
(e.g., normal human brain segmentation\textsuperscript{19}) and other
computer vision applications (e.g., human gait
detection\textsuperscript{20}). In the context of neuropathology, random
forest-based paradigms have been employed in the delineation of multiple
sclerosis lesions\textsuperscript{21}, stroke
lesions\textsuperscript{22}, and brain tumors\textsuperscript{23--26}
for both single and multi-modal acquisition protocols. Of note, these
latter random forest approaches for brain tumor segmentation have
performed well in recent international competitions
\replaced[id=dt]{established in response to the lack of objective comparisons between segmentation algorithms (i.e.,  the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in 2012}{
In response to the lack of objective comparisons between segmentation algorithms, the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in 2012 [@Menze:2015aa] and has continued every year since under the auspices of the MICCAI conference.}\textsuperscript{27}.

Random forests are conceptually straightforward\textsuperscript{18}.
They consist of ensembles of decision trees that are built from training
data. Once constructed, data to be classified is ``pushed'' through each
decision tree resulting in a single classification ``vote'' per tree.
These votes are then used for regression or classification of the data.
Although decision trees had been extensively studied
\deleted[id=dt]{previously}, the success of employing collections of
such weak learners for boosting machine learning performance (e.g.,
AdaBoost\textsuperscript{28,29}) influenced the similarly sytled
conglomeration of decision trees into ``forests'' with randomized node
optimization\textsuperscript{30,31}. Finally,
Breiman\textsuperscript{18} improved accuracy by random sampling of
\deleted[id=dt]{the} training data (i.e., ``bagging'') resulting in the
current random forest technique \added[id=dt]{applied here}.

In this work, we develop a concatenated random forest framework with a
tailored contextual feature image set (both spatial and intensity-based)
for segmenting \replaced[id=dt]{WMHs}{white matter hyperintensities} in
\replaced[id=dt]{a arge TBI cohort}{traumatic brain injury cohorts}.
Additionally, the entire framework is provided publicly through the
well-known open-source \added[id=lw]{Advanced Normalization Tools}
(ANTs)\footnote{\url{https://github.com/stnava/ANTs}} and
ANTsR\footnote{\url{https://github.com/stnava/ANTsR}} toolkits.
\replaced[id=dt]{Further motivating this research is the availability of several large publicly available  imaging data sets that permits testing reproducibility of this automate routine for WMH segmentation and quantification.}{Further motivating this research of this work is
the public availability of the imaging data thus permitting full reproducibility
of the results reported and discussed.}

@@

MR images utilized for this initial report were acquired from 24
individuals scanned at a single site involved in the Chronic Effects of
Neurotrauma Consortium's (CENC) observational study
(\added[id=nt,remark=NT: reference needed?]{see [Walker], this issue}).
Briefly, participants were Operation Enduring Freedom/ Operation Iraqi
Freedom /Operation New dawn (OEF/OIF/OND) era service members and
veterans between the ages of 18--60 who have experienced combat
situations and are between X and X (period of time) from deployment.
\added[id=js,remark={JS: Data needed.}]{The feature images consisted of subjects aged $X.X \pm X$ years (range X – X years), XX percent male, with at least one or more potential concussive events (mean = X, SD = X, range X-X).}

@@

Crucial to these supervised segmentation approaches are the creation and
selection of ``features'' as input in conjunction with
\replaced[id=lw,remark=NT: I replaced ground-truth since it is problematic anyway]
{expertly identified structures of interest}{the ground-truth} for model
construction. For the targeted application in this work (i.e., WMHs),
regression/classification are performed at the voxelwise level. In other
words, each voxel within the region of interest is sent through the
ensemble of decision trees and receives a set of classification votes
from each tree thus permitting a regression or classification solution.
Since this procedure is performed at the voxelwise level, intensity
information alone is insufficient for good segmentation performance due
to the lack of spatial context. For example, as pointed out
in\textsuperscript{32}, higher intensities can be found at the
periventricular caps in normal subjects which often confounds automated
lesion detection algorithms. Other potential confounds include MR signal
inhomogeneity and noise. Therefore, even though machine learning and
pattern recognition techniques are extremely powerful and have
significant potential, just as crucial to outcome is the creative
construction and deployment of salient feature images which we detail
below.

@@

\input{featureImagesTable.tex}

@@

\begin{figure}[htbp]
\centering
\includegraphics{Figures/featureImages.png}
\caption{Representation of Stage 1 feature images for subject 01C1019.
The FLAIR, T1-, and T2-weighted images are
\added[id=am,remark=NT: no need for 12 dof for
intra-subject normalization]{rigidly} pre-aligned\textsuperscript{33} to
the space of the T1 image. The three modality images are then
preprocessed (N4 bias correction\textsuperscript{34} and adaptive
denoising\textsuperscript{35}) followed by application of standard ANTs
brain extraction and \(n\)-tissue segmentation protocols using the MMRR
symmetric template and corresponding priors\textsuperscript{36} applied
to the T1 image. The feature images are then generated for voxelwise
input to the RF model which results in the voting maps illustrated on
the right. This gives a probabilistic classification of tissue type. Not
shown are the probability and voting images for the brain stem and
cerebellum.}
\end{figure}

As mentioned previously, input for each subject comprises FLAIR, T1-,
and T2-weighted acquisitions. The FLAIR and T2 images are rigidly
registered to the T1 image using the open-source Advanced Normalization
Tools (ANTs)\textsuperscript{33}. The aligned images are then
preprocessed using the denoising algorithm of\textsuperscript{35}
followed by N4 bias correction\textsuperscript{34} which are then
normalized to the intensity range \([0,1]\). Although we could have used
an alternative intensity standardization algorithm
(e.g.,\textsuperscript{37}), we found that a simple linear rescaling
produced better results similar
\added[id=dt]{to previous work}\textsuperscript{26}.

\begin{figure}[htbp]
\centering
\includegraphics{Figures/sampleResults.png}
\caption{Sample FLAIR acquisition image slices showing both manual and
random forest segmentations for both stages obtained during the
leave-one-out evaluation. Manual segmentations were performed by one of
the authors and provided the ground truth WMH labels for training the
random forest models.}
\end{figure}

The T1 image is then processed via the ANTs brain extraction and normal
tissue segmentation (i.e., the pathology is discounted which is a fairly
safe assumption for the T1 image) protocols as part of the cortical
thickness pipeline described in\textsuperscript{36}.
\added[id=hl]{This protocol involves preprocessing using N4 bias correction followed by a template-based strategy for brain extraction.  Once the
brain has been extracted, we apply a Bayesian-based segmentation algorithm
with a template-based prior probability strategy to segment the
parenchymal tissue types.  The result is} a mask delineating the brain
parenchyma and probabilistic estimates of the
\replaced[id=dt]{CSF}{cerebrospinal fluid (csf)}, gray matter, white
matter, deep gray matter, brain stem, and cerebellum. These provide the
expertly annotated labels for the first six tissue labels given above.
The \replaced[id=dt]{WMHs}{white matter hyperintensities} were manually
identified by one of the authors (J. R. S.) using the ITK-SNAP
tool\textsuperscript{38}. Segmentation is performed using the ANTs
Atropos tool\textsuperscript{39} and multi-model optimal symmetric
shape/intensity templates\textsuperscript{26} created from the public
MMRR data set\textsuperscript{40} (cf Figure 3).

\begin{figure}[htbp]
\centering
\includegraphics{Figures/MMRR.png}
\caption{Canonical views of the mutlivariate, bilaterally symmetric
template constructed from the MMRR data set\textsuperscript{40} (only
shown are the FLAIR, T1, and T2 modalities--- the components relevant
for this work). Template construction is detailed
in\textsuperscript{26}. These images are important for specific
intensity-based features.}
\end{figure}

To model the intensity information the first set of images simply
includes the preprocessed and normalized intensity FLAIR, T1, and T2
image voxel values. We also calculate a set of neighborhood statistics
(mean, standard deviation, and skewness) feature images using a
\added[id=am]{Manhattan} radius of one voxel
\added[id=nt]{given the typical size
of individual WMHs}. For each of the normalized images, we calculate the
difference in intensities with the corresponding warped template
component. Previous success in the international brain tumor
segmentation competition\textsuperscript{27} was based on an important
set of intensity features that were created from multi-modal templates
mentioned previously\textsuperscript{26}
\added[id=nt]{and listed in Table 1}. We employ the same strategy here.
For example, the template difference feature image for the FLAIR image,
\(S_{FLAIR}\) is calculated as:

@@

where \(\phi_b: S \leftrightarrow \underset{b}{\leftrightsquigarrow} T\)
is the transform which maps from the individual subject space to the
template space and \(T_{FLAIR}\) is the FLAIR template component. Also,
to take advantage of the
\added[id=hl,remark=NT:  we are using the gross bilateral symmetry of the brain in an analogous sense that if we know the left eye location and color then we can accurately predict the location and color of the right eye even though a particular face might not be exactly symmetric]{gross}
bilateral symmetry of the normal brain (in terms of both shape and
intensity), and the fact that
\replaced[id=bt]{WMHs do not generally manifest symmetrically across hemispheres}{the presence of WMHs violates that assumption},
we use the symmetric templates to compute the contralateral intensity
differences as an additional intensity feature. For the FLAIR component,
this contralateral difference image is calculated from

@@

The segmentation probability images described above are used as feature
images to provide a spatial context for the random forest model
prediction step. Additional spatial contextual feature images include
the distance maps\textsuperscript{41} based on the csf, gray matter, and
deep gray matter images. These latter images are intended to help
distinguish white matter hyperintensities from false positives induced
by the partial voluming at the gray/white matter interface. A third set
of images are based on the voxel location within the space of the
template. The T1 image of the subject is registered to the T1 template
component using a B-spline variant\textsuperscript{42} of the well-known
ANTs Symmetric Normalization (SyN) algorithm\textsuperscript{43}. Since
the inverse transform is also derived as part of the registration
process, we can warp the voxel index locations back to the space of the
individual subject
\replaced[id=nt]{which motivates similar work by others}{
Note that this is similar in motivation to the work
of}\textsuperscript{44}. However, this previous work lacks the
normalization to the standard coordinate system provided by the template
to dramatically improve spatial specificity across all subjects.

@@

In previous brain tumor segmentation work\textsuperscript{26}, it was
demonstrated that a concatenated supervised approach, whereby the
prediction output from the first random forest model serves as partial
input for a second random forest model, can significantly improve
segmentation performance. We do the same thing for the work described
here
\added[id=lw]{where we employ two stacked random forests (or two "stages")}.
The Stage 1 feature images of the training data (as described
previously) are used to construct the Stage 1 model. The training data
Stage 1 features are then used to produce the voxelwise
\added[id=lw]{"voting maps" (i.e., the classification count of each decision tree for each tissue label)}
via the Stage 1 random forest model. All the Stage 1 features plus the
Stage 1 voting maps are used as input to the Stage 2 model. In addition,
we use the Stage 1 voting maps as tissue priors
\added[id=lw]{(i.e., probabilistic estimates of the tissue spatial locations)}
for a second application of the Atropos maximum aposteriori algorithm
with an additional Markov Random Field spatial prior
(MAP-MRF)\textsuperscript{39}. However, for the second stage we use all
three aligned preprocessed images for a multivariate segmentation. The
resulting seven posterior probability images constitute a third
additional feature image set for Stage 2.

@@

As pointed out in a recent comprehensive multiple sclerosis lesion
segmentation review\textsuperscript{45}, although the number of
algorithms reported in the literature is quite extensive, there were
only four publicly available segmentation algorithms at the time of
writing \added[id=am]{this article}.
\replaced[id=dt]{In contrast to the current work,}{of which} none are
based on supervised learning. As we did for our brain tumor segmentation
algorithm\textsuperscript{26}, all of the code described in this work is
publicly available through the open-source ANTs/ANTsR toolkits. Through
ANTsR (an add-on toolkit which, in part, bridges ANTs and the R
statistical project) we use the \texttt{randomForest}
package\textsuperscript{46} using the default settings with 2000 trees
per model and 500 randomly selected samples per label per image.
\added[id=bt]{Note that we saw little variation in performance when these parameters were changed (i.e. up to 1000 random samples and as little as 1000 trees) which is consistent with our previous experience.}

In addition, similar to our previous offering,\footnote{\url{https://github.com/ntustison/ANTsAndArboles}}
we plan on creating a self-encapsulated example to showcase the proposed
methodology. The fact that the data will also be made available through
the
\replaced[id=lw]{Federal Interagency Traumatic Brain Injury Research (FITBIR)}{FITBIR}
repository along with the manual labelings will facilitate
reproducibility on the part of the reader as well as any interest in
extending the proposed framework to other data sets.

@@

  \begin{itemize}
  \tightlist
  \item
    sequester the current subject and corresponding feature images.
  \item
    construct the Stage 1 random forest model from the remaining 23
    subjects.
  \item
    apply the Stage 1 random forest model to the feature images of the
    23 training subjects.
  \item
    the previous step produces the Stage 1 voting maps for all seven
    labels.
  \item
    for each of the 23 subjects, perform a Bayesian-based segmentation
    with an MRF spatial prior using the seven voting maps are used as
    additional tissue priors.
  \item
    construct the Stage 2 random forest model from all the Stage 1
    feature images, seven voting maps, and seven posterior probability
    maps from the previous step.
  \item
    send the sequestered subject through the random forest models for
    both stages.
  \item
    compare the final results with the manually-defined white matter
    hyperintensity regions.
  \end{itemize}
\end{itemize}

@@

After performing the leave-one-out
evaluation\deleted[id=dt]{described at the end of the previous section},
we calculated the \texttt{MeanDecreaseAccuracy} feature values for each
of the 24 subjects \(\times\) 2 models per subject \(=48\) total models.
This measure (per feature, per model) is calculated during the
out-of-bag phase of the random forest model construction and quantifies
the decrease in prediction accuracy from omitting the specified feature.
In other words, this quantity helps determine the importance of a
particular feature and, although we save such efforts for future work,
this information provides us with guidance for future feature pruning
and/or additions.

\begin{figure}[htbp]
\centering
\includegraphics{Figures/averageLeaveOneOutStage1.png}
\caption{Average \texttt{MeanDecreaseAccuracy} plots generated from the
creation of all 24 random forest models for Stage 1 during the
leave-one-out evaluation. These plots are useful in providing a
quantitative assessment of the predictive importance of each feature.
\added[id=nt]{Features are ranked in descending order of importance.}
The horizontal error bars provide the \(95^{th}\) percentile
\deleted[id=bt]{(i.e., $1.96 \times \sigma$)} and illustrate the
stability of the feature importance across the leave-one-out models.
\added[id=nt]{At this initial stage only 31 features images are used.}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{Figures/averageLeaveOneOutStage2.png}
\caption{Average \texttt{MeanDecreaseAccuracy} plots generated from the
creation of all 24 random forest models for Stage 2 during the
leave-one-out evaluation. These plots are useful in providing a
quantitative assessment of the predictive importance of each feature.
\added[id=nt]{Features are ranked in descending order of importance.}
The horizontal error bars provide the \(95^{th}\) percentile
\deleted[id=bt]{(i.e., $1.96 \times \sigma$)} and illustrate the
stability of the feature importance across the leave-one-out models.
\added[id=nt]{We augment the 31 feature images from the first stage by adding an additional
7 voting maps and 7 segmentation posteriors from application of the Bayesian-based
segmentation for a total of 45 images for the second stage.}}
\end{figure}

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
    \setmainfont{Georgia}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

\usepackage{booktabs}
\usepackage[final]{changes}
\usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
\linespread{2}
\usepackage[compact]{titlesec}
\usepackage{enumitem}
\usepackage{tikz}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
\setlist{nolistsep}
\titlespacing{\section}{2pt}{*0}{*0}
\titlespacing{\subsection}{2pt}{*0}{*0}
\titlespacing{\subsubsection}{2pt}{*0}{*0}
\setlength{\parskip}{3pt}
\setremarkmarkup{(#2)}
\definechangesauthor[name={Nick Tustison}, color=red]{nt}
\definechangesauthor[name={James Stone}, color=magenta]{js}
\definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
\definechangesauthor[name={Andy Mayer}, color=green]{am}
\definechangesauthor[name={Harvey Levin}, color=blue]{hl}
\definechangesauthor[name={Brian Taylor}, color=purple]{bt}
\definechangesauthor[name={David Tate}, color=orange]{dt}

@@

Machine learning and pattern recognition techniques have seen increased
application for various medical image analysis workflows (see, for
example, the annual Workshop on Machine Learning in Medical Imaging held
in conjunction with the Medical Image Computing and Computer-Aided
Intervention (MICCAI) international meeting). Popular techniques such as
support vector machines and neural networks have been applied
successfully to clinically relevant imaging tasks such as supervised
image segmentation (e.g.,\textsuperscript{11}) and diagnostic prediction
(e.g.,\textsuperscript{12,13}). Facilitating the current employment of
such techniques are the number of available imaging data
sets\textsuperscript{14} and the public availability of data science
packages such as SciPy\textsuperscript{15} and the R project for
statistical computing\textsuperscript{16} and their associated
extensions.

The random forests framework\textsuperscript{17} is a popular machine
learning technique that has demonstrated significant utility for
supervised segmentation tasks (e.g., normal human brain
segmentation\textsuperscript{18}) and other computer vision applications
(e.g., human gait detection\textsuperscript{19}). In the context of
neuropathology, random forest-based paradigms have been employed in the
delineation of multiple sclerosis lesions\textsuperscript{20}, stroke
lesions\textsuperscript{21}, and brain tumors\textsuperscript{22--25}
for both single and multi-modal acquisition protocols. Of note, these
latter random forest approaches for brain tumor segmentation have
performed well in recent international competitions established in
response to the lack of objective comparisons between segmentation
algorithms (i.e., the Multimodal Brain Tumor Segmentation (BRATS)
challenge was initiated in 2012\textsuperscript{26}.

Random forests are conceptually straightforward\textsuperscript{17}.
They consist of ensembles of decision trees that are built from training
data. Once constructed, data to be classified is ``pushed'' through each
decision tree resulting in a single classification ``vote'' per tree.
These votes are then used for regression or classification of the data.
Although decision trees had been extensively studied, the success of
employing collections of such weak learners for boosting machine
learning performance (e.g., AdaBoost\textsuperscript{27,28}) influenced
the similarly sytled conglomeration of decision trees into ``forests''
with randomized node optimization\textsuperscript{29,30}. Finally,
Breiman\textsuperscript{17} improved accuracy by random sampling of
training data (i.e., ``bagging'') resulting in the current random forest
technique applied here.

In this work, we develop a concatenated random forest framework with a
tailored contextual feature image set (both spatial and intensity-based)
for segmenting WMHs in a large TBI cohort. Additionally, the entire
framework is provided publicly through the well-known open-source
Advanced Normalization Tools (ANTs)\footnote{\url{https://github.com/stnava/ANTs}}
and ANTsR\footnote{\url{https://github.com/stnava/ANTsR}} toolkits.
Further motivating this research is the availability of several large
publicly available imaging data sets that permits testing
reproducibility of this automated routine for WMH segmentation and
quantification.

@@

MR images utilized for this initial report were acquired from a single
scanner involved in the Chronic Effects of Neurotrauma Consortium's
(CENC) observational study (see Walker et al., this issue). Briefly,
participants were Operation Iraqi Freedom/Operation Enduring Freedom
(OIF/OEF) era Service Members and Veterans between the ages of 18-60
years with prior combat exposure and deployment(s). The feature images
consisted of 26 subjects aged 39.6 ± 8.1 years (range 28 -- 58 years).
Within this cohort, 24 (92\%) were considered positive for TBI based
upon the potential concussive events (PCE) interview process described
in detail in Walker et al., this issue). Each of the participants that
were selected from the larger cohort of participants in this study
demonstrated at least one white matter hyperintensity (but as many as
20) on FLAIR imaging.

@@

Crucial to these supervised segmentation approaches are the creation and
selection of ``features'' as input in conjunction with expertly
identified structures of interest for model construction. For the
targeted application in this work (i.e., WMHs), tissue classification is
performed at the voxelwise level. In other words, each voxel within the
region of interest is sent through the ensemble of decision trees and
receives a set of classification votes from each tree thus permitting a
regression or classification solution. Since this procedure is performed
at the voxelwise level, intensity information alone is insufficient for
good segmentation performance due to the lack of spatial context. For
example, as pointed out in\textsuperscript{31}, higher intensities can
be found at the periventricular caps in normal subjects which often
confounds automated lesion detection algorithms. Other potential
confounds include MR signal inhomogeneity and noise. Therefore, even
though machine learning and pattern recognition techniques are extremely
powerful and have significant potential, just as crucial to outcome is
the creative construction and deployment of salient feature images which
we detail below.

@@

@@

As mentioned previously, input for each subject comprises FLAIR, T1-,
and T2-weighted acquisitions. The FLAIR and T2 images are rigidly
registered to the T1 image using the open-source Advanced Normalization
Tools (ANTs)\textsuperscript{32}. The aligned images are then
preprocessed using the denoising algorithm of\textsuperscript{33}
followed by N4 bias correction\textsuperscript{34} which are then
normalized to the intensity range \([0,1]\). Although we could have used
an alternative intensity standardization algorithm
(e.g.,\textsuperscript{35}), we found that a simple linear rescaling
produced better results similar to previous work\textsuperscript{25}.

The T1 image is then processed via the ANTs brain extraction and normal
tissue segmentation pipelines\textsuperscript{36}. Any effect of
pathology is minimal because WMHs are not commonly seen in T1-weighted
MRI. This protocol involves preprocessing using N4 bias correction
followed by a template-based strategy for brain extraction. Once the
brain has been extracted, we apply a Bayesian-based segmentation
algorithm with a template-based prior probability strategy to segment
the parenchymal tissue types. The result is a mask delineating the brain
parenchyma and probabilistic estimates of the CSF, gray matter, white
matter, deep gray matter, brain stem, and cerebellum. These provide the
expertly annotated labels for the first six tissue labels given above.
The WMHs were manually identified by one of the authors (J. R. S.) using
the ITK-SNAP tool\textsuperscript{37}. Segmentation is performed using
the ANTs Atropos tool\textsuperscript{38} and multi-model optimal
symmetric shape/intensity templates\textsuperscript{25} created from the
public MMRR data set\textsuperscript{39} (cf Figure 3).

To model the intensity information the first set of images simply
includes the preprocessed and normalized intensity FLAIR, T1, and T2
image voxel values. We also calculate a set of neighborhood statistics
(mean, standard deviation, and skewness) feature images using a
Manhattan radius of one voxel given the typical size of individual WMHs.
For each of the normalized images, we calculate the difference in
intensities with the corresponding warped template component. Previous
success in the international brain tumor segmentation
competition\textsuperscript{26} was based on an important set of
intensity features that were created from multi-modal templates
mentioned previously\textsuperscript{25} and listed in Table 1. We
employ the same strategy here. For example, the template difference
feature image for the FLAIR image, \(S_{FLAIR}\) is calculated as:

@@

where

\[\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T\]

is the transform which maps from the individual subject space to the
template space and \(T_{FLAIR}\) is the FLAIR template component. Also,
to take advantage of the gross bilateral symmetry of the normal brain
(in terms of both shape and intensity), and the fact that WMHs do not
generally manifest symmetrically across hemispheres, we use the
symmetric templates to compute the contralateral intensity differences
as an additional intensity feature. For the FLAIR component, this
contralateral difference image is calculated from

@@

The segmentation probability images described above are used as feature
images to provide a spatial context for the random forest model
prediction step. Additional spatial contextual feature images include
the distance maps\textsuperscript{40} based on the csf, gray matter, and
deep gray matter images. These latter images are intended to help
distinguish white matter hyperintensities from false positives induced
by the partial voluming at the gray/white matter interface. A third set
of images are based on the voxel location within the space of the
template. The T1 image of the subject is registered to the T1 template
component using a B-spline variant\textsuperscript{41} of the well-known
ANTs Symmetric Normalization (SyN) algorithm\textsuperscript{42}. Since
the inverse transform is also derived as part of the registration
process, we can warp the voxel index locations back to the space of the
individual subject which motivates similar work by
others\textsuperscript{43}. However, this previous work lacks the
normalization to the standard coordinate system provided by the template
to dramatically improve spatial specificity across all subjects.

@@

In previous brain tumor segmentation work\textsuperscript{25}, it was
demonstrated that a concatenated supervised approach, whereby the
prediction output from the first random forest model serves as partial
input for a second random forest model, can significantly improve
segmentation performance. We do the same thing for the work described
here where we employ two stacked random forests (or two ``stages''). The
Stage 1 feature images of the training data (as described previously)
are used to construct the Stage 1 model. The training data Stage 1
features are then used to produce the voxelwise ``voting maps'' (i.e.,
the classification count of each decision tree for each tissue label)
via the Stage 1 random forest model. All the Stage 1 features plus the
Stage 1 voting maps are used as input to the Stage 2 model. In addition,
we use the Stage 1 voting maps as tissue priors (i.e., probabilistic
estimates of the tissue spatial locations) for a second application of
the Atropos maximum aposteriori algorithm with an additional Markov
Random Field spatial prior (MAP-MRF)\textsuperscript{38}. However, for
the second stage we use all three aligned preprocessed images for a
multivariate segmentation. The resulting seven posterior probability
images constitute a third additional feature image set for Stage 2.

@@

As pointed out in a recent comprehensive multiple sclerosis lesion
segmentation review\textsuperscript{44}, although the number of
algorithms reported in the literature is quite extensive, there were
only four publicly available segmentation algorithms at the time of
writing this article. In contrast to the current work, none are based on
supervised learning. As we did for our brain tumor segmentation
algorithm\textsuperscript{25}, all of the code described in this work is
publicly available through the open-source ANTs/ANTsR toolkits. Through
ANTsR (an add-on toolkit which, in part, bridges ANTs and the R
statistical project) we use the \emph{randomForest}
package\textsuperscript{45} using the default settings with 2000 trees
per model and 500 randomly selected samples per label per image. Note
that we saw little variation in performance when these parameters were
changed (i.e.~up to 1000 random samples and as little as 1000 trees)
which is consistent with our previous experience.

In addition, similar to our previous offering,\footnote{\url{https://github.com/ntustison/ANTsAndArboles}}
we plan on creating a self-encapsulated example to showcase the proposed
methodology. The fact that the data will also be made available through
the Federal Interagency Traumatic Brain Injury Research (FITBIR)
repository along with the manual labelings will facilitate
reproducibility on the part of the reader as well as any interest in
extending the proposed framework to other data sets.

@@

  \begin{itemize}
  \tightlist
  \item
    sequester the current subject and corresponding feature images.
  \item
    construct the Stage 1 random forest model from the remaining 23
    subjects.
  \item
    apply the Stage 1 random forest model to the feature images of the
    23 training subjects.
  \item
    the previous step produces the Stage 1 voting maps for all seven
    labels.
  \item
    for each of the 23 subjects, perform a Bayesian-based segmentation
    with an MRF spatial prior using the seven voting maps as additional
    tissue priors.
  \item
    construct the Stage 2 random forest model from all the Stage 1
    feature images, seven voting maps, and seven posterior probability
    maps from the previous step.
  \item
    send the sequestered subject through the random forest models for
    both stages.
  \item
    compare the final results with the manually-defined white matter
    hyperintensity regions.
  \end{itemize}
\end{itemize}

@@

After performing the leave-one-out evaluation, we calculated the
\emph{MeanDecreaseAccuracy} feature values for each of the 24 subjects
\(\times\) 2 models per subject \(=48\) total models. This measure (per
feature, per model) is calculated during the out-of-bag phase of the
random forest model construction and quantifies the decrease in
prediction accuracy from omitting the specified feature. In other words,
this quantity helps determine the importance of a particular feature
and, although we save such efforts for future work, this information
provides us with guidance for future feature pruning and/or additions.

The resulting rankings for both Stages are given in Figures 4 and 5
where the values for the separate stages are averaged over the entire
corresponding model set. In addition, we track the variance for each
feature over all models to illustrate the stability of the chosen
features during the evaluation. This latter information is illustrated
as horizontal errors bars providing the \(95^{th}\) percentile Note that
the reader can cross reference Table 1 for identifying corresponding
feature types and names.

One can also use these measurements as a type of sanity check. For
example, from the Stage 1 plot, one can see that the
\emph{MeanDecreaseAccuracy} values for the location indices in the
anterior-posterior direction (i.e., \emph{TemplateIndicesWarped1}) are
greater than those for either the inferior-superior (i.e.,
\emph{TemplateIndicesWarped2}) or the left-right (i.e.,
\emph{TemplateIndicesWarped0}) directions in the space of the symmetric
template.

Additionally, it is interesting to note some of the other top performing
features for Stage 1. The contralateral difference FLAIR image is highly
discriminative over the set of evaluation random forest models (see
Figure 6). This accords with the known clinical relevance of FLAIR
images for identifying white matter hyperintensities and the fact that
such pathology does not typically manifest symmetrically in both
hemispheres. Interestingly, the posterior maps for the deep gray matter
are extremely important for accurate white matter hyperintensity
segmentation. Perhaps the spatial specification of deep gray matter aids
in the removal of false positives. Inspection of the bottom of the plots
demonstrates the lack of discriminating features associated with the T1
image which is also well-known in the clinical literature.
